---
title: "Understanding a linear model"
date: "last-modified"
echo: true 
warning: false 
message: false
---


```{r}
#| echo: FALSE
library(tidyverse)
library(emmeans)
library(gt)
library(kableExtra)
```

In the previous chapter, we showed how the mean captures the effects of the first levels of each factor included in the model. As illustrated below, block1 and geng1 are missing from the model matrix. The reason for this omission is to avoid linearly dependent columns—columns that can be expressed as linear combinations of others. If the design matrix $X$ were not full rank, it would not be invertible, and the system would have no unique solution.

```{r}
data <- read.csv("../data/example_1.csv") |>
  mutate(gen = as.factor(gen), block = as.factor(block))

n <- 12
n_b <- 3
n_g <- 4

X <- model.matrix(yield ~ 1 + block + gen, data = data)
y <- matrix(data[, "yield"])
print(X)
```


Thus, we do not obtain any coefficient regarding these two levels

```{r}
Xty <- t(X) %*% y
XtX <- t(X) %*% X
XtX_inv <- solve(XtX)
beta <- XtX_inv %*% Xty
beta
```

We also covered how to obtain these values 'manually' by adding to the estimated mean 
the proportioinal effect of the other factors

```{r}
# Betas
beta_geng1 <- beta[1] + sum(beta[2:3])/3
beta_block1 <- beta[1] + sum(beta[4:6])/4 
```

Which mathematically represent the following operations

$$BLUE(geng1) = \mu + \frac{1}{3}\hat\beta_{block2} + \frac{1}{3}\hat\beta_{block3} = 7.75 - \frac{1.65}{3} - \frac{2.10}{3} = 6.5$$

$$BLUE(block1) = \mu + \frac{1}{4}\hat\beta_{geng2} + \frac{1}{4}\hat\beta_{geng3} + \frac{1}{4}\hat\beta_{geng4} = 7.75 + \frac{1.10}{4} + \frac{0.10}{4} + \frac{1.80}{4} = 8.5$$

Additionally, we can also express the estimated coefficients in the desired units rather than based
on the intercept. So, $BLUE(geng2)$ would be

```{r}
beta_block2 <- beta[1] + sum(beta[4:6])/4 + beta[2]
```

Which equals

$$BLUE(geng2) = \mu + \frac{1}{3}\hat\beta_{block2} + \frac{1}{3}\hat\beta_{block3} + \hat\beta_{geng2}$$
Then, for the remaining levels

```{r}
beta_block3 <- beta[1] + sum(beta[4:6])/4 + beta[3]

beta_geng2 <- beta[1] + sum(beta[2:3])/3  + beta[4]
beta_geng3 <- beta[1] + sum(beta[2:3])/3  + beta[5]
beta_geng4 <- beta[1] + sum(beta[2:3])/3  + beta[6]

blues <- matrix(c(beta_block1, beta_block2, beta_block3, beta_geng1, beta_geng2, beta_geng3, beta_geng4), ncol = 1)
rownames(blues) <- c('block1', 'block2', 'block3', 'geng1', 'geng2', 'geng3', 'geng4')
blues
```

Now, imagine we have multiple factors with tens of levels each one. Doing this manually would be
tedious. So let's explore an easier way to solve this. Specifically, we will make use of 
the $L$ matrix to compute all desired values simultaneously, so that

$$BLUE(g) = L\hat\beta$$

The $L$ matrix represents the levels we want to compute. For example, if we want to 
extract the BLUEs of the genotypes, we should include the intercept, the fraction 
of each block level that we have to add to the mean, and the levels that we want to 
compute from the genotypes. Notice how L has as many rows as levels are in the factor to be solved.

```{r}
L <- cbind(
  matrix(1, nrow = n_g, ncol = 1), # Intercept
  matrix(1 / n_b, nrow = n_g, ncol = n_b - 1), # Average block
  matrix(rbind(0, diag(nrow = n_g - 1)), nrow = n_g, ncol = n_g - 1)
)
L
```

Now we can simply multiply this matrix times the estimated coefficients

```{r}
BLUEs <- L %*% beta
BLUEs
```

The way the BLUEs are calculated equals the operations shown above, where

$$
\small
\boldsymbol{BLUEs(gen)} = \boldsymbol{L}\,\boldsymbol{\hat\beta} =
\begin{bmatrix}
\overset{\text{Intercept}}{1} & 
\overset{\text{block2}}{0.33} & 
\overset{\text{block3}}{0.33} & 
\overset{\text{geng2}}{0} & 
\overset{\text{geng3}}{0} & 
\overset{\text{geng4}}{0} \\
1 & 0.33 & 0.33 & 1 & 0 & 0 \\ 
1 & 0.33 & 0.33 & 0 & 1 & 0 \\ 
1 & 0.33 & 0.33 & 0 & 0 & 1 \\ 
\end{bmatrix}
\,
\overset{\text{mean}}{
\begin{bmatrix}
7.75 \\
-1.65 \\
-2.10 \\
1.10 \\
0.10 \\
1.80
\end{bmatrix}} = 
\begin{bmatrix}
6.5 \\
7.6 \\
6.6 \\
8.3
\end{bmatrix}
$$

So that

$$
BLUE(gen)[1,1] = BLUE(geng1) = 1*7.75 + \frac{1}{3}*-1.65 + \frac{1}{3}*-2.10 = 6.5
$$



Similarly, to obtain the BLUEs of each block we would do

```{r}
L2 <- cbind(
  matrix(1, nrow = n_b, ncol = 1), # Intercept
  matrix(rbind(0, diag(nrow = n_b - 1)), nrow = n_b, ncol = n_b - 1), # Average block
  matrix(1 / n_g, nrow = n_b, ncol = n_g - 1)
)
L2
```

```{r}
BLUEs2 <- L2 %*% beta
BLUEs2
```

You can check that the results are the same as computing them manually

```{r}
cbind(blues, rbind(BLUEs2, BLUEs))
```

Now let's compute the variance and the SE of the BLUEs. Now we now that $BLUEs = L\hat\beta$, 
thus

$$Var(L\hat\beta) = LVar(\hat\beta) L^T = LC_{11} L^T$$

We will discuss more in detail
about the Coefficient Matrix once we start digging into Linear Mix Models. 
For now, we will just say that $C_{ii}$ corresponds to $Var(\hat\beta)$, so 

```{r}
# Compute the errors to compute sigma^2 
y_hat <- X %*% beta
errors <- y - y_hat
SSE <- sum(errors^2)
sigma_2 <- SSE / (n - 6)
C_11 <- solve(t(X)%*%X) * sigma_2

C_11
```

Then the Variance and the SE of the BLUEs are

```{r}
var_BLUEs <- L %*% C_11 %*% t(L)
var_BLUEs |> round(4)

# SE BLUEs
se_BLUEs <- sqrt(diag(var_BLUEs))
se_BLUEs

data.frame(
  gen = levels(data$gen),
  BLUEs = BLUEs,
  var_BLUEs = diag(var_BLUEs),
  se_BLUEs = se_BLUEs
)
```


We can also use `lme` package to fit the model and extract the coefficients 
and `emmeans` to retrieve the BLUEs. This is useful since emmeans returns 
the $1$ and $C_{11}$ matrices, so you don't have to build them yourself. 

```{r}
mod <- lm(formula = yield ~ 1 + block + gen, data = data)
beta_mod <- coef(mod)
mod

mm <- emmeans(mod, ~gen)
mm
L_emm <- mm@linfct
C_11_emm <- mm@V

BLUE_mod <- L_emm %*% beta_mod
var_BLUEs_emm <- L_emm %*% C_11_emm %*% t(L_emm)
sqrt(diag(var_BLUEs_emm))
```


Besides that, it allows us to compute the estimates and the variances of the differences
between levels, such as genotypes. 

```{r}
diff_mm <- emmeans(mod, pairwise ~ gen)
diff_mm$contrasts

L_diff <- diff_mm$contrasts@linfct
L_diff # See how L now links more than 1 genotype per row

BLUEs_diff <- L_diff %*% beta_mod
rownames(BLUEs_diff) <- diff_mm$contrasts@levels$contrast

rownames(L_diff) <- rownames(BLUEs_diff)
var_diff <- L_diff %*% C_11_emm %*% t(L_diff)
var_diff |> round(4) # Notice how now there are covariances between the levels, since they share one of the genotypes in common
SE <- sqrt(diag(var_diff))

avg_diff <- mean(SE)
avg_diff
```















