---
title: "Untitled"
editor: visual
---

```{r}
#| echo: false
#| warning: false
rm(list = ls())
library(viridis)
library(sommer)
library(ggforce)
library(AGHmatrix)
library(StageWise)
library(CVXR)
library(asreml)
library(tidyverse)
```

# Theoretical Background

Selecting individuals, both for product development and parental selection, based
on multiple traits is a rather common practice in plant breeding. That
selection is typically done based on an index that summarises all available information 
for a genotype into a single value. 

A selection index for $t$ standardized traits is defined as 

$$
\mathbf{I} = \sum_{k = 1}^tb_ky_k = \mathbf{b}'\mathbf{y}
$$

where $b$ represents the index coeffficients. Typically, it is assumed that 
$\sigma_I^2 = 1$, thus

$$
\operatorname{var}(\mathbf{b'y}) = \mathbf{b'}\operatorname{var}(\mathbf{y})\mathbf{b} = \mathbf{b'Pb}
$$

Under the multitrait framework, $\operatorname{var}(\mathbf{y})$ refers to the 
phenotypic variance covariance matrix between traits so that 
$P_{jk} = \operatorname{cov}(y_j, y_k)$. The response to truncation selection  on the 
index follows the breeders equation so that

$$
R_{g_j} = \frac{i}{\sigma_I}\operatorname{cov}(g_j, I) = i\sum_{k = 1}^t\operatorname{cov}(g_j, y_k)b_k = i\sum_{k = 1}^t\Gamma_{jk} b_k
$$

It is important to notice that under BLUP selection, the formulation 
of the index changes to 

$$
\mathbf{I} = \sum_{k = 1}^tb_k\hat{g}_k = \mathbf{b}'\mathbf{\hat{g}}
$$

Therefore, the variance of the index now represents $\sigma_I^2 = 1 = \mathbf{b'}\operatorname{var}(\mathbf{\hat{g}})\mathbf{b}$
so that $P_{jk} = \operatorname{cov}(\hat{g}_j, \hat{g}_k)$. Additionally, 
the response would be defined as 

$$
R_{g_j} = i\sum_{k = 1}^t\operatorname{cov}(g_j, \hat{g}_k)b_k = i\sum_{k = 1}^t\Gamma_{jk} b_k
$$

In this particular case, given the properties of the BLUP 
$\operatorname{var}(\hat{\mathbf{g}}) = \operatorname{cov}(\mathbf{g}, \hat{\mathbf{g}})$, 
$\Gamma_{jk} = \operatorname{cov}(g_j, \hat{g}_k) = P_{jk}$

An standardize response for each trait can be obtained by mutiplying by the 
inverse of a diagonal matrix $H$ of the genetic standard deviations

$$
x = i\mathbf{H}^{-1}\mathbf{\Gamma b}
$$

From where we can solve for the index to obtain that

$$
\mathbf{b} = i^{-1}\mathbf{\Gamma}^{-1} \mathbf{H x} 
$$

# Desired Gain Index

The desired gains coefficients represent the vector of standardize response that 
we would like to achive for each of the traits of interest. However, normally these
gains cannot be achieved, since the resopnse must satisfy that $\mathbf{b'Pb} = 1$
and therefore the following equation

$$
\begin{align}
i^{-1} \, \mathbf{x}^\top \mathbf{H} \mathbf{\Gamma}^{-1} \mathbf{P} \mathbf{\Gamma}^{-1} \mathbf{H} \mathbf{x} \, i^{-1} &= 1 \\
i^{-2} \, \mathbf{x}^\top \mathbf{H} \mathbf{\Gamma}^{-1} \mathbf{P} \mathbf{\Gamma}^{-1} \mathbf{H} \mathbf{x} &= 1 \\
\mathbf{x}^\top \mathbf{H} \mathbf{\Gamma}^{-1} \mathbf{P} \mathbf{\Gamma}^{-1} \mathbf{H} \mathbf{x} &= i^2 \\
\mathbf{x}^\top \mathbf{Q} \mathbf{x} &= i^2
\end{align}
$$

The equation above $\mathbf{Q} = \mathbf{H} \mathbf{\Gamma}^{-1} \mathbf{P} \mathbf{\Gamma}^{-1} \mathbf{H}$ represents the matrix representation of an ellipsoide (high dimensional ellipse). 
The shape and surface of the ellipse (if only 2 traits are considered) is 
determined by the intensity of selection, the correlation between traits the heritability
and the accuracy of each one of them.

The response thus represents the intersection between a vector director (the desired vector)
and an ellipse (or ellipsoide). This, mathematically, is obtained as follows

First, the response is defined as the desired vector times a proportionality constant 
$t$ 

$$
\mathbf{x} = t\mathbf{d}
$$

Then, we substitute in the constraint equation and obtain

$$
(t\mathbf{d})'\mathbf{Q}t\mathbf{d} = i^2 \\
t^2\mathbf{d}'\mathbf{Qd} = i^2 \\
t = \frac{i}{\sqrt{\mathbf{d}'\mathbf{Qd}}}
$$

And therefore, the maximized response is given by $\mathbf{x} = \frac{i}{\sqrt{\mathbf{d}'\mathbf{Qd}}} \mathbf{d}$, 
which represents where does the vector of desired gains touch the surface of the ellipse. 

# Genetic Merit

A different approach is to specify the relative importance of genetic gains for 
each trait, which is known as the economic weights or merit coefficients.  The
merit function is defined as follows. 

$$
M = \sum_{k = 1}^t a_k(\frac{g_k}{\sigma_{gk}}) = \mathbf{a'H^{-1}g}
$$

Now, the merit coefficients measure not the genetic gains but its value. The goal is then 
to maximize the correlation between the merit or genotypic worth and the Index, 
agiain, subkject to $\sigma^2_I = 1$. Therefore, the quantity to be maximized is 

$$
\begin{align}
\operatorname{cov}(\mathbf{M, I}) 
&= \operatorname{cov}(\mathbf{a'H^{-1}g}, \mathbf{b'y}) \\
&= \mathbf{a'H^{-1}}\operatorname{cov}(\mathbf{g}, \mathbf{g})\mathbf{b}' \\
&= \mathbf{a'H^{-1}} \mathbf{\Gamma b'}
\end{align}
$$

Thus, the optimization problem can be defined as

$$
\begin{aligned}
\text{maximize} \quad & \mathbf{a'H^{-1} \Gamma b'} \\
\text{subject to} \quad & \mathbf{b' P b = 1}
\end{aligned}
$$

We can obtain the Lagrange and then solve for b by taking its partial derivative and 
equating to 0 so that

$$
\begin{aligned}
\mathcal{L}(\mathbf{b}, \lambda) 
&= \mathbf{a'H^{-1} \Gamma b'} - \frac{\lambda}{2} (\mathbf{b' P b} - 1) \\[2mm]
\frac{\partial \mathcal{L}}{\partial \mathbf{b}} 
&= \mathbf{\Gamma' H^{-1} a} - \lambda \mathbf{P b} = \mathbf{0} \\[1mm]
\mathbf{P b} 
&= \frac{1}{\lambda} \mathbf{\Gamma' H^{-1} a} \\[1mm]
\mathbf{b} 
&= \frac{1}{\lambda} \mathbf{P^{-1} \Gamma' H^{-1} a}
\end{aligned}
$$

If we substitute this $b$ values into the Index equation, we would obtain that

$$
I_{SH} = \mathbf{a'H^{-1}\Gamma P^{-1}y} = \frac{\mathbf{a'H^{-1}\hat{\mathbf{g}}}}{\lambda}
$$

Additionally, if we substitute in the constraint

$$
\begin{align*}
\mathbf{b' P b} &= 1 \\[1mm]
\mathbf{\frac{1}{\lambda} a' H^{-1} \Gamma P^{-1} P P^{-1} \Gamma H^{-1} a \frac{1}{\lambda}} &= 1 \\[1mm]
\frac{\mathbf{a' H^{-1} \Gamma P^{-1} \Gamma H^{-1} a}}{\lambda^2} &= 1 \\[1mm]
\frac{1}{\lambda} &= \frac{1}{\sqrt{\mathbf{a' Q^{-1} a}}}
\end{align*}
$$

We can now subsitute that into the coefficients 

$$
\mathbf{b} = \frac{1}{\lambda} \mathbf{P^{-1} \Gamma' H^{-1} a} 
= \frac{\mathbf{P^{-1} \Gamma' H^{-1} a}}{\sqrt{\mathbf{a' Q^{-1} a}}}
$$


# Exercise

We will use data from the UW Madison potato breeding program. The data
contains measurements for several traits for 176 individuals. We will run the analysis
only for Yield and average_GC, for which the phenotypic covariance is around -0.33

```{r}
traits <- c('yield', 'average_GC')
pheno <- read.csv('../data/example_2.csv') |> 
  na.omit() |> 
  select(uid, row, range, gen, traits) |> 
  mutate(
    across(all_of(traits), \(x) as.numeric(scale(x))), 
    across(c(gen, uid), as.factor)
  )
head(pheno)
```


To obtain $\mathbf{P}$ and $\mathbf{\Gamma}$ we need to fit the following 
multitrait model

$$
\mathbf{y} = \mathbf{\mu} + \mathbf{X \beta} + \mathbf{Z g} + \mathbf{\varepsilon}
$$

assuming that $\mathbf{g} \sim MVN(0, \mathbf{G})$, 
$\mathbf{\varepsilon} \sim MVN(0, \mathbf{R})$ and $\mathbf{y} \sim MVN(\mathbf{X \beta}, \mathbf{V})$, 
where $\mathbf{G} = \mathbf{I}_n \otimes \mathbf{\Sigma_g}$ and 
$\mathbf{R} = \mathbf{I}_{nt} \otimes \mathbf{\Sigma_e}$.


We will use asreml and sommer to estimates the variance-covariance matrices 
$\mathbf{\Sigma_g}$ and $\mathbf{\Sigma_e}$. For the genetic correlation we will 
assume and unstructure variance covariance matrix, meaning that the model will estimate
$n*(n+2)/2$ paremeters. In this case, it will estimate a variance for each of the traits plus 
a covaraince components between them. For the residuals, sommer (at least `mmes`)
does not allow to estimate covariances between the errors for the different traits. 
Thus, a diagonal matrix will be used. For asreml, an additional unstructured matrix
will be employed. 

highlight the importance of estimating the correlation 
between error amongst traits and how does the choice of that variance covariance matrix
determine the estimation of the additive correlation between traits. 

```{r}
mod <- asreml(
  fixed = cbind(yield, average_GC) ~ trait, 
  random = ~ us(trait):gen,
  residual = ~ id(units):us(trait),
  verbose = F, 
  data = pheno
)

sigma_g <- matrix(0, nrow = 2, ncol = 2)
sigma_g[upper.tri(sigma_g)] <- summary(mod)$varcomp[2, 1]
sigma_g[lower.tri(sigma_g, diag = T)] <- summary(mod)$varcomp[1:3, 1]

sigma_e <- matrix(0, nrow = 2, ncol = 2)
sigma_e[upper.tri(sigma_e)] <- summary(mod)$varcomp[6, 1]
sigma_e[lower.tri(sigma_e, diag = T)] <- summary(mod)$varcomp[5:7, 1]

dimnames(sigma_g) <- dimnames(sigma_e) <- list(traits, traits)
cov2cor(sigma_g)
cov2cor(sigma_e)
```

```{r}
#| warning: false

pheno_long <- pheno |>  
  pivot_longer(
    cols = traits,
    names_to = 'trait',
    values_to = 'value'
    ) |>
  arrange(trait) |>
  mutate(across(c(trait, gen, uid), as.factor))

mod_sommer <- mmes(
  fixed = value ~ trait,
  random = ~ vsm(usm(trait), ism(gen)),
  rcov = ~ vsm(dsm(trait), ism(units)),
  data = pheno_long,
  verbose = F
)
sommer_sigma_g <- cov2cor(mod_sommer$theta$`vsm(usm(trait), ism(gen`)
sommer_sigma_e <- cov2cor(mod_sommer$theta$units)
dimnames(sommer_sigma_g) <- dimnames(sommer_sigma_e) <- list(traits, traits)
cov2cor(sommer_sigma_g)
cov2cor(sommer_sigma_e)
```

Now we will reconstruct all the necessary matrices to show how to compute 
$\mathbf{P}$ and $\mathbf{\Gamma}$. Firstwe will compute $\mathbf{G}$, 
$\mathbf{R}$ and $\mathbf{V}$ and then used it to compute the Prediction Error
Variance (PEV) and the marginal variance of the BLUPs ($\operatorname{var}(\hat{\mathbf{g}})$)

```{r}
n <- n_distinct(pheno$gen)
t <- n_distinct(traits)
ids <- unique(pheno$gen)
names_G <- as.vector(outer(traits, ids, paste, sep = ":"))

y <- as.matrix(pheno_long[, 'value'], ncol = 1)
G <- diag(nrow = n) %x% sigma_g 
Z <- model.matrix(~ -1 + gen:trait, pheno_long)
X <- model.matrix(~ trait, pheno_long)
R <- diag(nrow(pheno)) %x% sigma_e 
V <- Z%*%G%*%t(Z) + R
V_inv <- solve(V)
P <- V_inv - V_inv %*%X%*%solve(t(X)%*%V_inv%*%X)%*%t(X)%*%V_inv

C11 <- t(X) %*% solve(R) %*% X 
C12 <- t(X) %*% solve(R) %*% Z
C21 <- t(Z) %*% solve(R) %*% X 
C22 <- t(Z) %*% solve(R) %*% Z + solve(G)

C <- rbind(
  cbind(C11, C12), 
  cbind(C21, C22)
)
C_inv <- solve(C)
rhs <- rbind(
  t(X) %*% solve(R) %*% y, 
  t(Z) %*% solve(R) %*% y
)
ans <- C_inv %*% rhs

PEV <- G - G%*%t(Z)%*%P%*%Z%*%G
var_uhat <- G - PEV
```


As mentioned above,$\mathbf{\Gamma}$ represents the genotypic variance covariance
matrix between traits (meaning, its a $t \times t$ matrix). If only random effects are
considered, it is computed  as described by  [Legarra 2016](https://www.sciencedirect.com/science/article/pii/S0040580915000830?via%3Dihub)
and [Endelman 2023](https://pubmed.ncbi.nlm.nih.gov/36949348/)

$$
E[V_y] = \overline{\operatorname{diag}(\mathbf{L})} - \overline{L_{..}}
$$

where $\mathbf{L}$ represents the marginal variance of BLUPs with genotypes nested
traits. However, right now the information in the marginal variance matrix is organized with 
traits nested genotypes, meaning that the diagonal is composed of $n$ $t \times t$ 
variance covariance matrices each genotype for the two traits 

$$
\operatorname{var}(\hat{\mathbf{g}}) = 
\begin{bmatrix}
M_{g_1} & \cdots & M_{g_1g_n} \\
\vdots & \ddots & \vdots \\
M_{g_ng_1} & \cdots & M_{g_n}
\end{bmatrix} = 
\begin{bmatrix}
\underbrace{
\begin{bmatrix}
\sigma^2_{t_1} & \sigma_{t_1 t_2} \\
\sigma_{t_2 t_1} & \sigma^2_{t_2}
\end{bmatrix}
}_{g_1}
& \cdots & M_{g_1g_n} \\
\vdots & \ddots & \vdots \\
M_{g_ng_1} & \cdots & 
\underbrace{
\begin{bmatrix}
\sigma^2_{t_1} & \sigma_{t_1 t_2} \\
\sigma_{t_2 t_1} & \sigma^2_{t_2}
\end{bmatrix}
}_{g_n}
\end{bmatrix}
$$


The reason why this happens is because of how we computed $\mathbf{G}$ 
taking into account the kronecker product properties

$$
\mathbf{G} = \mathbf{I_n} \otimes \mathbf{\Sigma_g}
=
\mathbf{I_n} \otimes
\begin{bmatrix}
\Sigma_{g,11} & \Sigma_{g,12} \\
\Sigma_{g,21} & \Sigma_{g,22} 
\end{bmatrix}
=
\begin{bmatrix}
\mathbf{\Sigma_g} & 0 & \cdots & 0 \\
0 & \mathbf{\Sigma_g} & \cdots & 0 \\
\vdots & \vdots & \ddots & \vdots \\
0 & 0 & \cdots & \mathbf{\Sigma_g}
\end{bmatrix}
$$

The same structure is thus found for $\operatorname{var}(\mathbf{\hat{g}})$ 
In summary

```{r}
names <- as.vector(outer(traits, unique(pheno$gen), paste, sep = ":"))
dimnames(G) <- dimnames(var_uhat) <- list(names, names)
var_uhat[1:2, 1:2]
var_uhat[3:4, 3:4]
```

Therefore, we need to rearrange the matrix to compute the population covariance

```{r}
nt <- nrow(var_uhat)
idx <- split(1:nt, f = rep(1:t, times = n))
B <- matrix(0, ncol = t, nrow = t)
for (i in 1:t) {
  for (j in i:t) {
    idxi <- idx[[i]]
    idxj <- idx[[j]]
    L <- var_uhat[idxi, idxj]
    B[i, j] <- B[j, i] <- mean(diag(L)) - mean(L)
  }
}
dimnames(B) <- list(traits, traits)
P <- Gamma <- B # P equals Gamma since we are working under BLUP
H <- diag(sqrt(diag(sigma_g)))
dimnames(H) <- list(traits, traits)
Gamma
```

As stated in StageWise documentation, the quadratic matrix can be computed 
as  $\mathbf{Q} = \mathbf{U}'\mathbf{U}$, where $\mathbf{U} = \mathbf{R} \mathbf{\Gamma}^{-1} \mathbf{H}$ 
and $\mathbf{R}$ is the upper triangular cholesky factor of $\mathbf{P}$. Therefore

```{r}
U <- chol(P) %*% solve(Gamma) %*% H 
Q <- crossprod(U)
Q
```

## Desired

Now we only need to define our desired gains to obtain the maximal response given 
our ellipse. Let's imagine we want to simultaneously advance both traits despite
its negative correlation

```{r}
#| warning: false
desired <- c(1, 0)
intensity <- 1
x.opt <- desired * intensity/sqrt(crossprod(desired, Q%*%desired)) # apply dimensionality constraint
b <- (solve(Gamma) %*% H %*% x.opt)/intensity #substitute response to solve for coefficients
b <- b/sqrt(sum(b^2)) # normalize vector
desired_results <- data.frame(
  traits = traits, 
  response = x.opt, 
  coefficients = b
)
desired_results
```

We can additionally plot the selection tradeoff. The dot represent the response, 
while the dashed line represents the direction of the desired gain vector. 

```{r}
eg <- eigen(Q[1:2, 1:2])
lens <- intensity / sqrt(eg$values)
angle <- atan(eg$vectors[2, 2] / eg$vectors[1, 2])
p1 <- ggplot() +
  geom_ellipse(
    aes(x0 = 0, y0 = 0, a = lens[2], b = lens[1], angle = angle)
  ) +
  coord_fixed() +
  theme_bw() +
  xlab(traits[1]) +
  ylab(traits[2]) +
  geom_segment(
    aes(
      x = 0,
      y = 0,
      xend = x.opt[[1]],
      yend = x.opt[[2]]
    ),
    col = "red",
    lty = 2
  ) +
  geom_point(
    aes(
      x = x.opt[[1]],
      y = x.opt[[2]]
    ),
    col = "red"
  )

p1
```


## Merit

To show the differences between the two approaches, we will use the same 
coefficients but following the Merit function. Therefore we will obtain the 
maximized response and from there the coefficients

```{r}
#| warning: false
merit <- c(1, 0)
x <- Variable(t)
constraints <- list(quad_form(x, Q) <= intensity^2)
v <- matrix(merit, nrow = 1)
objective <- Maximize(v %*% x)
problem <- Problem(objective, constraints)
result <- solve(problem)
x <- as.numeric(result$getValue(x))
b <- (solve(Gamma) %*% H %*% x) / intensity
b <- b / sqrt(sum(b^2))
merit_results <- data.frame(
  traits = traits,
  response = x,
  coefficients = b
)
merit_results

x2 <- merit * intensity / sqrt(as.numeric(crossprod(merit, Q %*% merit)))
p2 <- ggplot() +
  geom_ellipse(aes(x0 = 0, y0 = 0, a = lens[2], b = lens[1], angle = angle)) +
  coord_fixed() +
  theme_bw() +
  xlab(traits[1]) +
  ylab(traits[2]) +
  geom_segment(
    aes(x = 0, y = 0, xend = x2[[1]], yend = x2[[2]]),
    col = "red",
    lty = 2
  ) +
  geom_point(aes(x = x2[[1]], y = x2[[2]]), col = "red") +
  geom_segment(
    aes(x = 0, y = 0, xend = x[[1]], yend = x[[2]]),
    col = "blue"
  ) +
  geom_point(aes(x = x[[1]], y = x[[2]]), col = "blue")

p2
```

# Selection

Based on those estimates, we can extract the BLUPs, build the Indeces and
perform truncation selection based on it. See how depending on the index 
the individuals selected vary.

```{r}
blups <- as.matrix(ans[grep('gen', rownames(ans), value = T), ]) |> 
  as.data.frame() |> 
  rownames_to_column('gen') |> 
  mutate(
    gen = gsub('^gen', '', gen),
    gen = gsub('trait', '', gen)) |> 
  separate(gen, into = c('gen', 'trait'), sep = ':') |> 
  arrange(trait, gen) |> 
  rename(blup = V1)

I_desired <- list()
for (i in seq_along(unique(blups$trait))){
  trait_tmp <- unique(blups$trait)[i]
  b_tmp <- desired_results |> filter(traits == trait_tmp) |>  pull(coefficients)
  I_desired[[i]] <- blups |>  filter(trait == trait_tmp) |> pull(blup) * b_tmp
}
I_desired <- Reduce('+', I_desired)

I_merit <- list()
for (i in seq_along(unique(blups$trait))){
  trait_tmp <- unique(blups$trait)[i]
  b_tmp <- merit_results |> filter(traits == trait_tmp) |>  pull(coefficients)
  I_merit[[i]] <- blups |>  filter(trait == trait_tmp) |> pull(blup) * b_tmp
}
I_merit <- Reduce('+', I_merit)

plot_df <- blups |> 
  pivot_wider(names_from = trait, values_from = blup) 
plot_df <- cbind(plot_df, I_desired, I_merit) |> 
  pivot_longer(col = c(I_desired, I_merit), names_to = 'Index', values_to = 'value')

ggplot(plot_df, aes(x = average_GC, y = yield, color = value)) + 
  geom_point() +
  scale_color_viridis_c(option = "plasma") +
  theme_bw() + 
  facet_grid(.~Index)
```



















